---
title: "Get started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The code behind our [methodology report](https://stapm.gitlab.io/model-inputs/utility_report/qaly_estimation_report.pdf) is in [this repo](https://gitlab.com/stapm/model-inputs/utility_report).   

We show the use of the `qalyr` functions in the workflows below. At the moment, these workflows are accessible only to the project team - who are added to the STAPM Gitlab organisation.    

- [Estimation of health state utility values](https://gitlab.com/stapm/model-inputs/utility_value_estimation)  

## Basic functionality

### Load and merge data 

The data is stored in `D:/QALYs/Raw/`. The following code will read, clean, filter, and combine the data. 

```{r readdata, eval = F}

library(qalyr)

# Merge inpatient data and HODAR data and collapse to survey level data 
data <- qalyr::merge_data(
  inpatient_data = qalyr::read_inpatient_data(),
  survey_data = qalyr::read_survey_data(),
  lkup = qalyr::lkup)

# With this data we can calculate the condition-specific HSUVs, and the mean age and proportion of 
# males by condition. These are both used as inputs to calculate the age-sex-condition-specific HSUVs 

# Condition-specific utilities
cond_utils <- qalyr::CalcUtils(data, lkup = qalyr::lkup)

# Mean age and proportion of males 
mean_age_sex <- qalyr::MeanAge_Sex(data, lkup = qalyr::lkup)

# age-sex-condition-specific utilities 
adj_cond_utils <- qalyr::Adj_cond_utils(Age_Sex_data = mean_age_sex,
                                 Condition_utilities = cond_utils,
                                 GenPopUtil = qalyr::GenPopUtil)


```


